Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-07-12 AnalyticStructure_main.tex
update

\title{Locality and Computational Reliability of Linear Response Calculations for Molecular Systems}% Localization properties, observable character and computational reliability of the linear response in open systems}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}
\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), INAC, CEA-UGA, 17 Av. des Martyrs,
38054 Grenoble, France}
\date{\today}

\begin{abstract}
By performing a critical analysis of the fundamental equations of linear-response  (LR) formalisms in molecules, 
we explore the interplay between locality of the response density operator and  numerical convergence of LR-related quantities.
We show that the first ionization potential of the system behaves as a threshold energy: for frequencies below this value, it is possible to parametrize the response density by employing localized states only. Above this threshold, such a locality property cannot be achieved. 
Such considerations may be transposed in terms of the molecule's excited states. We show that not all the system's excitations can be considered on equal footing.
There is a discrete sector of excitations that can be parametrized by observable, localized states, which can be computationally expressed with high precision, provided an adequate level of completeness. 
We present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  
The remaining excitation modes belong to a continuum spectrum that, on the contrary, is not directly associated to observable properties and can only be effectively represented in a given computational setup.
Such considerations are important not only for reproducibility of the results among different computer codes 
employing diverse formalisms, but also in view of providing a deeper understanding on the impact of models' 
approximations on the scientific outcomes of the simulation.

% ---
% We discuss these aspects in the context of linear response treatment of molecular systems. 
% By analysing simple case-studies with a highly complete numerical formalism, 
% we explore the interplay between locality of the perturbed wavefunctions and 
% numerical convergence of observable quantities. This will enable us to draw some considerations on the analytic structure of the linear susceptivity for open systems and on the potential convergence issues that might arise in a given computational setup.

% We believe that such discussion could provide useful considerations and perspectives for the design of a effective approaches for the study of molecular excitations in realistic conditions.

% We analyze how the fundamental equations which govern the linear response properties of a quantum system are influenced by the presence of open boundaries.
% In particular we focus our attention on linear-response (LR)-TDDFT calculation of molecular systems.
% Above this energy threshold, the excitation spectrum may be separated in two sectors: one can still be described in terms of observable excitation 
% with discrete energies, that can be captured within computational setups tailored for localized states. 
% This result is a inherent behavior of the system's Liouvillian, and does not depend on computational treatment of the 
% unoccupied subspace.
\end{abstract}

@@

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems, 
in particular referring to LR-TDDFT calculation. We will investigate the impact that 
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the computational evaluation of LR quantities.
In order to do this, we perform analytic considerations on the equations beyond LR treatment.
Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - 
from others that would \emph{explicitly} depend of the employed computational treatment. 
In particular, we will discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects.
% In order to do this, we perform analytic considerations on the equations beyond LR treatment,
% which will help us identifying how to distinguish -- \textit{a priori} -- quantities that are susceptible to be 
% compared among different treatments, as they are \emph{intrinsically} associated to localized objects,
% from quantities that \emph{explicitly} depend of the the employed computational treatment.

@@

% Moreover, wheareas for Ground-State calculations the quantities are relatively easy to identify and extract,
% TD-DFT and more generally, linear-response calculation present a further difficulty, which is the problem
% of \emph{numerical} convergence. In other terms, beyond ground state, 
% the absence of tools like the variational theorem imposes us to focus on \emph{well-defined} quantites,
% and to study the conditions under which numerical convergence of the results can be achieved.

% The subject of the present investigation concerns with the study of isolated molecular-like systems subjected to a time-dependent perturbation. % in the linear response regime.
% This problem can be tackled within the contest of the time-dependent density functional theory (TD-DFT) \cite{runge1984,onida2002}, that allows one to establish the time-dependent 
% Kohn-Sham (KS) equations which determine the evolution of the one-particle properties of the system with respect to time. 
% 
% We focus our attention to the linear response regime (LR) in which the theory can be conveniently expressed in the frequency domain and the time-dependent KS equation are reformulated 
% as a quantum Liouville equation for the response density operator. Different computational approaches have been developed in literature \cite{baroni2008,hubener2014,brabec2015}, principally 
% with the aim of producing efficient algorithms for the evaluation of the physical observables. These methods differ both for the procedure adopted to solve the Liouville equation 
% (explicit diagonalization, self-consistent Sternheimer equation, Lanczos chain-like procedure, etc.) and both for the choice of the basis set employed to perform the calculations. 
% Nonetheless, in all the cases one is a dealing with a finite basis set that, unavoidably, gives rise to a limited computational domain. 
% It is then meaningful to try to understand under 
% which extent the computational-induced truncation of the natural open boundaries condition could produce observable effects on the physical quantities computed in the LR. 
% 
% A preliminary understanding of this problem can be achieved thanks to the results reported in \cite{boffi2016}. The authors of this paper discuss the effects of a finite basis set on the 
% occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, and the independence of its properties (\emph{e.g.} energy) 
% with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
% able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
% true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
% Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.  
% 
% In this paper we aim to reach a deeper understanding of the possible effects of a finite computational basis set on the LR regime in open systems. To achieve this task we try to
% reformulate the relation between localization properties and basis independence for the main actors of the LR formalism. We present a discussion based on a double level analysis. First, 
% we propose a localization argument directly based on the inspection of the response density operator. Then, we move to a deeper level and investigate the properties of the susceptibility 
% functional, which represents the fundamental object in the contest of the linear response. Pursuing this approach we are naturally led to investigate the localization properties of excitation 
% operators, that in turn emerge as the resonant channel of the spectral decomposition of the susceptibility. In this way we establish a direct relation between the concept of basis independence 
% and the analytic structure of the linear response susceptibility.

% First of all it is useful to discuss the property of the ground state under this perspective. 
% Indeed, it is well known that the stability of the ground state and the localized nature of a molecular-like system ensures that the KS description of occupied states is represented 
% by a discrete set of wave functions that approaches to zero far apart from a compact region around the system's atoms. This behavior indicates that a (finite) basis set will provide 
% a reliable description of the ground state properties if it is able to sample a computational domain which contains the natural sizes of the occupied orbitals \cite{boffi2016}.
% The same arguments cannot be rigidly extended to the case in which an external time-dependent field starts acting on the system. Indeed, the perturbation drives the system out 
% from the equilibrium and the electrons sample a wider region of the state space, that will contains orbitals of both negative (bound) and positive (unbound) energies.

%\section{Localization properties of the linear response actors in open systems}
\subsection{Long range behavior of the eigenvalue solution of the Schr\"odinger equation}
\label{SEopenSystem}

@@

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}, where the authors analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, 
and the independence of its energy with respect to the basis set. 
% It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
% able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
% true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, it is clearly shown that the energies of the unbound orbitals are not stable with respect to parameters like the size of the computational domain.

\section{Fluctuation states. A localization argument for the response density}
It is evident that the arguments presented above have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a 
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems that are electronically stable (i.e. no anionic or metastable configurations),
such quantity is localized, in the sense of the above section. Stated otherwise, 
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

Let us now inspect the case of LR calculations.
%Linear response can be effectively applied to the analysis of the electronic excitations in open systems.
In this framework one consider the unperturbed Hamiltonian $\hnot$ and assume that 
the GS density $\dmnot$ is accessible and expressed in a given computational treatment. 
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a 
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system. 
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

%The main objective of this section is to investigate the structure of the building blocks needed to express the response density operator. In particular, we are interested in clarifying 
%the behavior of the main actors of the LR in the long range, \emph{i.e.} outside from the support domain of the potential.

% As a preliminary step for the forthcoming discussion we present a simple argument that allows us to determine the asymptotic character of the eigenvalue solution of the Schr\"odinger equation (SE) 
% $\op H\ket{\psi}=\eps\ket{\psi}$. The main aims of this analysis, apart from reproducing the well known behavior of the wave functions, is to introduce some tools and setting up notations and 
% terminology that will we widely used in the rest of the paper. 
% 
% The argument proceeds by splitting the Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term. Since we are dealing with molecular-like systems $\op V$ has a compact support when 
% projected in the $\r$-representation. We can formally solve the SE as an inhomogeneous Helmholtz equation by writing
% \be\lb{SEHelmholtz1}
% \ket{\psi} = \op G_H(\eps)\ket{\op V\psi}\;,
% \ee
% where we have introduced the Green function of the Helmholtz operator $\op G_H(\eps) = (\eps-\op T)^{-1}$. We observe that, due to the properties of $\op V$, the \emph{source term} $\ket{\op V\psi}$ 
% of equation \eqref{SEHelmholtz1} is localized in space and, for open boundary conditions in three dimensions, the kernel of $\op G_H(\eps)$ in the coordinate representation reads: 
% \be\lb{HelmholtzKernelDef1}
% \bra{\r} \op G_H(\eps) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
% \frac{e^{-\alpha|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; \eps  < 0\; \\ 
% \frac{e^{\ii \alpha |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; \eps \geq 0
% \end{cases} \;,
% \ee
% where $\alpha = \sqrt{2|\eps|}$. 
% This representation of the eigenvalues solution of the SE evidences that the value $\eps=0$ represents the threshold level that determines the asymptotic character of the wave functions. 
% Indeed, for $\eps > 0$,  $\op G_H(\eps)$ behaves as a spherical wave and its convolution with the source term  gives raise to a $\psi(\r)$ delocalized in the whole accessible space. Instead, 
% for values of $\eps < 0$, the kernel contains an exponential damping factor and the associated wave functions exhibit a bound-state behavior in the long range. Finally, we observe that the 
% square-integrability and the continuity of the wave function and its derivatives poses further constrains responsible for the discretization of the energy spectrum for bound-states.  
% 
% The different asymptotic behavior of the wave functions has a deep impact on the possible computational induced error obtained when this quantities are evaluated using a (finite) computational
% setup.

\label{FluctuationState}

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;.
\ee
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$ encodes the response of the $\dm$-dependent potential to a modification of the density operator.
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$.
An operator $\op O_\perp$ with this feature satisfy the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\op O_\perp \equiv 
\dmnot\op O \op Q_0 + \op Q_0 \op O\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Thus the condition $\dm'_\Phi = \dm_{\perp}$ holds.

We can therefore introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states (FS) are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p} 
$ as a source term.

It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that governs the localization properties of the FS.
As a \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is localized, \textit{i.e.} its evaluation can be
restricted on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
%in the $\r$-representation. 
Even when such locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of 
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range, 
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
%\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
% is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
% solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line) 
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest 
box.}
\end{figure}

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} = 
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
% A direct confirmation of the localization character of fluctuation states at $\omega=0$ can be achieved through a ground-state calculation. Indeed, $\ket{f_p(\omega)}$ are nothing but the 
% perturbed KS eigenstates projected out from the domain of $\dmnot$, so for a constant perturbation $\op\Phi_s$ it turns out that $\ket{f_p(0)} = \op Q_0\ket{\psi_p^{\Phi_s}}$. The perturbed KS 
% states $\ket{\psi_p^{\Phi_s}}$ can be computed (beyond LR) by performing a DFT ground-state computation in which the constant perturbing field is added to the external potential and the 
% spatial extension of the fluctuation states can be directly measured.
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response 
density in terms of FS implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibits the locality properties introduced for the potentials $ \op V$, $\op \Phi$ and $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.

>>>>

\title{Locality and Computational Reliability of Linear Response Calculations for Molecular Systems}
\author{Marco D'Alessandro}
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}
\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), INAC, CEA-UGA, 17 Av. des Martyrs,
38054 Grenoble, France}
\date{\today}

\begin{abstract}
By performing a critical analysis of the fundamental equations of linear-response  (LR) formalisms in molecules, 
we explore the interplay between locality of the response density operator and  numerical convergence of LR-related quantities.
We show that the first ionization potential of the system behaves as a threshold energy: for frequencies below this value, it is possible to parametrize the response density by employing localized states only. Above this threshold, such a locality property cannot be achieved. 
Such considerations may be transposed in terms of the molecule's excited states. We show that not all the system's excitations can be considered on equal footing.
There is a discrete sector of excitations that can be parametrized by observable, localized states, which can be computationally expressed with high precision, provided an adequate level of completeness. 
We present indicators that can help to quantify such potential observable property of an excitation, that can be evaluated 
in any discretization scheme.  
The remaining excitation modes belong to a continuum spectrum that, on the contrary, is not directly associated to observable properties and can only be effectively represented in a given computational setup.
Such considerations are important not only for reproducibility of the results among different computer codes 
employing diverse formalisms, but also in view of providing a deeper understanding on the impact of models' 
approximations on the scientific outcomes of the simulation.
\end{abstract}

@@

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems, 
in particular referring to LR-TDDFT calculation. We will investigate the impact that 
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the computational evaluation of LR quantities.
In order to do this, we perform analytic considerations on the equations beyond LR treatment.
Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - 
from others that would \emph{explicitly} depend of the employed computational treatment. 
In particular, we will discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects.

@@

\subsection{Long range behavior of the eigenvalue solution of the Schr\"odinger equation}
\label{SEopenSystem}

@@

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}, where the authors analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, 
and the independence of its energy with respect to the basis set. 
Furthermore, it is clearly shown that the energies of the unbound orbitals are not stable with respect to parameters like the size of the computational domain.

\section{Fluctuation states. A localization argument for the response density}
\label{FluctuationState}
It is evident that the arguments presented above have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a 
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems that are electronically stable (i.e. no anionic or metastable configurations),
such quantity is localized, in the sense of the above section. Stated otherwise, 
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

Let us now inspect the case of LR calculations.
In this framework one consider the unperturbed Hamiltonian $\hnot$ and assume that 
the GS density $\dmnot$ is accessible and expressed in a given computational treatment. 
The linear response formalism allows us to evaluate the modification of the expectation value of a generic observable induced by a 
time (or frequency)-dependent perturbing field $\op\Phi(\omega)$ acting on the system. 
This quantity, written in the frequency domain, can be expressed through the evaluation of the \emph{linear response functional}
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}_\Phi(\omega) = \trace{\dm'_\Phi(\omega)\op O} \;,
\ee
where $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} under inspection 
and we have introduced the \emph{response density operator} $\dm'_\Phi(\omega)$ that codifies the modification of $\dmnot$ induced by the perturbation $\op\Phi(\omega)$.

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;.
\ee
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part $\Liouvnot$ plus a coupling term $ \coupl$. The action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$ encodes the response of the $\dm$-dependent potential to a modification of the density operator.
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidences that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$.
An operator $\op O_\perp$ with this feature satisfy the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\op O_\perp \equiv 
\dmnot\op O \op Q_0 + \op Q_0 \op O\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
In other terms, the condition $\dm'_\Phi = \dm'_{\Phi\,\perp}$ holds.

We can therefore introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states (FS) are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p} 
$ as a source term.

It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that governs the localization properties of the FS.
As a \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is localized, \textit{i.e.} its evaluation can be
restricted on a bounded real-space domain.
Even when such locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of 
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range, 
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line) 
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest 
box.}
\end{figure}

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} = 
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response 
density in terms of FS implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
Let us suppose that also the observable of interest $\op O$ exhibits the locality properties introduced for the potentials $ \op V$, $\op \Phi$ and $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.

We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enables us to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
